Investigations of thermal reactionomes reveal stress tolerance vs. resistance in

### 2014-08-19 doc/ApTranscriptome_manuscript.Rmd
some changes

---
title: Investigations of thermal reactionomes reveals stress tolerance vs. resistance
  warm and cold-climate ant species
author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
  James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
  Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli
  \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1}
  Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics
  Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences,
  North Carolina State University \\ \textsuperscript{4} Center for Macroecology,
  Evolution and Climate, University of Copenhagen}
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

\linenumbers
\tableofcontents

@@

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the evolution of thermal tolerance. We characterized the reactionomes of two common ant species in eastern U.S. forests: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. Interestingly, the warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change to thermal extremes than did *A. picea*. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. *A. carolinensis* shut down more genes than *A. picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A. picea*. These included genes functionally involved in metabolism, cellular organization and repair \todo{Sara - where did the 'repair' term come from?}. These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress, namely *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

# Significance Statement

**120-word-maximum statement about the significance of their research paper written at a level understandable to an undergraduate-educated scientist outside their field of specialty**

@@

The patterns of gene expression between species differed substantially, with more genes down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Fig. 1, 3). The results of the gene set enrichment analysis give an indication of what might drive this difference. The genes that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly species, shuts-down metabolism to a lesser extent than *A. picea* and so must rely on other mechanisms such as behavior (e.g. retreat into its nest) or proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*. Our results are at least congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. ...Sara - something about enzyme efficiency tradeoffs???

```{r max_exp_PDF, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
# reshape data
Ap.df$Species <- ifelse(Ap.df$colony == "ApVT", "A. picea", "A. carolinensis")

@@

print(maxexplot)
```

@@

@@

Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2011]. Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012].

>>>>

---
title: Investigations of thermal reactionomes reveal stress tolerance vs. resistance in
  warm and cold-climate ant species \\
author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
  James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
  Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli
  \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1}
  Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics
  Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences,
  North Carolina State University \\ \textsuperscript{4} Center for Macroecology,
  Evolution and Climate, University of Copenhagen}
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

\linenumbers

@@

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the evolution of thermal tolerance. We characterized the reactionomes of two common ant species in eastern U.S. forests: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. Interestingly, the warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change to thermal extremes than did *A. picea*. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. *A. carolinensis* shut down more genes than *A. picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A. picea*. These included genes functionally involved in metabolism, cellular organization, glycoprotein synthesis and DNA repair. These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress. Namely *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

The patterns of gene expression between species differed substantially, with more genes down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Fig. 3). The results of the gene set enrichment analysis give an indication of what might drive this difference. The genes that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly species, shuts-down metabolism to a lesser extent than *A. picea* and so must rely on other mechanisms such as behavior (e.g. retreat into its nest) or proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*. Our results are at least congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. ...Sara - something about enzyme efficiency tradeoffs???

```{r max_exp_PDF, eval = FALSE,aniopts= echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
# reshape data
Ap.df$Species <- ifelse(Ap.df$colony == "ApVT", "A. picea", "A. carolinensis")

@@

print(maxexplot)

####################################
### CHANGE TO REACTION NORM FOR HSPs 
####################################

```

@@

Thermal safety margins...

@@

Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010]. Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012].
